---
title: "info201Final"
output: html_document
date: "2023-12-01"
---

# START OF INFO 201 PROJECT 
## Downloading Data
```{r}
library(tidyverse)

gdp <- read.csv("gdp.csv")
global_education <- read.csv("global_education.csv")
```

## Merging Data
```{r}
gdp_global_education <- left_join(gdp, global_education, by = c("Country.Territory" = "Countries.and.areas"))
```

## Load Merged Data
```{r}
ls(gdp_global_education)
```

## Find the completion rate in males and females
```{r}
completionRate <- subset(gdp_global_education, select =c("Country.Territory", "Completion_Rate_Primary_Male", "Completion_Rate_Primary_Female"))

completionRate <- na.omit(completionRate)

completionRate
```


## Find the youth 15-24 literacy rate in males and females 
```{r}
youthLiteracyRate <- subset(gdp_global_education, select =c("Country.Territory", "Youth_15_24_Literacy_Rate_Female", "Youth_15_24_Literacy_Rate_Male"))

youthLiteracyRate <- na.omit(completionRate)

youthLiteracyRate
```


## Find gross primary enrollment in education 
```{r}
grossPrimaryEnrollment <- subset(gdp_global_education, select =c("Country.Territory", "Gross_Primary_Education_Enrollment"))

grossPrimaryEnrollment <- na.omit(grossPrimaryEnrollment)

grossPrimaryEnrollment
```

## Find unemployment rates
```{r}
unemploymentRates <- subset(gdp_global_education, select =c("Country.Territory", "Unemployment_Rate"))

unemploymentRates <- na.omit(unemploymentRates)

unemploymentRates
```

## COMPLETE MERGED DATASET WITH ALL VARIABLES WE NEED
```{r}
FINALgdp_global_education <- subset(gdp_global_education, select =c("Country.Territory", "Completion_Rate_Primary_Male", "Completion_Rate_Primary_Female", "Youth_15_24_Literacy_Rate_Female", "Youth_15_24_Literacy_Rate_Male", "Gross_Primary_Education_Enrollment", "Unemployment_Rate"  ))

FINALgdp_global_education <- na.omit(FINALgdp_global_education)

FINALgdp_global_education
```

## find income classes. make incomce classes and sort each country into its own bin for gdpp

