---
title: "info201Final"
output: html_document
date: "2023-12-01"
---

# START OF INFO 201 PROJECT 
# Downloading Data
```{r}
library(tidyverse)

gdp <- read.csv("gdp.csv")
global_education <- read.csv("global_education.csv")
```

# Data Joining
```{r}
gdp_global_education <- left_join(gdp, global_education, by = c("Country.Territory" = "Countries.and.areas"))
ls(gdp_global_education)
```

# Data Cleaning and Pre-processing
## If we need it: Find the completion rate in males and females
```{r}
completionRate <- subset(gdp_global_education, select =c("Country.Territory", "Completion_Rate_Primary_Male", "Completion_Rate_Primary_Female"))

completionRate <- na.omit(completionRate)
```


## If we need it: Find the youth 15-24 literacy rate in males and females 
```{r}
youthLiteracyRate <- subset(gdp_global_education, select =c("Country.Territory", "Youth_15_24_Literacy_Rate_Female", "Youth_15_24_Literacy_Rate_Male"))

youthLiteracyRate <- na.omit(completionRate)
```


## If we need it: Find gross primary enrollment in education 
```{r}
grossPrimaryEnrollment <- subset(gdp_global_education, select =c("Country.Territory", "Gross_Primary_Education_Enrollment"))

grossPrimaryEnrollment <- na.omit(grossPrimaryEnrollment)
```

## If we need it: Find unemployment rates
```{r}
unemploymentRates <- subset(gdp_global_education, select =c("Country.Territory", "Unemployment_Rate"))

unemploymentRates <- na.omit(unemploymentRates)
```

## MERGED DATASET WITH ALL VARIABLES WE NEED
```{r}
FINALgdp_global_education <- subset(gdp_global_education, select =c("Country.Territory", "IMF", "Completion_Rate_Primary_Male", "Completion_Rate_Primary_Female", "Unemployment_Rate", "OOSR_Primary_Age_Male", "OOSR_Primary_Age_Female"))

## gets rid of na values
FINALgdp_global_education <- na.omit(FINALgdp_global_education)

## gets rid of all 0 values in gdp (IMF)
FINALgdp_global_education <- FINALgdp_global_education[FINALgdp_global_education$IMF != 0, ]
```

### IMF = GDP per capita estimated by the International Monetary Fund
```{r}
FINALgdp_global_education$IMF <- as.numeric(FINALgdp_global_education$IMF)

min_value <- min(FINALgdp_global_education$IMF)
max_value <- max(FINALgdp_global_education$IMF)

## to see min and max
print(paste("Minimum value:", min_value))
print(paste("Maximum value:", max_value))
```

## Adding a categorical variable to the dataset: make different income classes, and sort each country into its own bin.
```{r}
FINALgdp_global_education <- mutate(FINALgdp_global_education,
                       Rank_of_GDP_per_Capita = case_when(
                           IMF >= 0 & IMF <= 1000 ~ "Low",
                           IMF >= 1000 & IMF <= 5000 ~ "Mid",
                           IMF >= 5001 & IMF <= 25000 ~ "High",
                           TRUE ~ NA_character_  # Catch-all for other cases
                       ))

rank_counts <- FINALgdp_global_education %>%
  count(Rank_of_GDP_per_Capita)

# Display the counts
print(rank_counts)

```

### FINAL DATASET
```{r}
ls(FINALgdp_global_education)
```

## Plot each GDP bin with its corresponding completion rate for primaryrate(female)
```{r}
library(ggplot2)

# Filter out rows where Completion_Rate_Primary_Female is not equal to 0
filtered_data <- FINALgdp_global_education %>%
  filter(Completion_Rate_Primary_Female != 0)

# Create a boxplot with adjustments using the filtered data
gdp_comp_rate_female <- ggplot(filtered_data, aes(x = factor(Rank_of_GDP_per_Capita, levels = c("Low", "Mid", "High")), y = Completion_Rate_Primary_Female)) +
  geom_boxplot(fill = "skyblue", color = "blue", alpha = 0.7, width = 0.5) +
  labs(title = "GDP per Capita Rank vs. Completion Rate for Primary Education (Female)",
       x = "Rank of GDP per Capita",
       y = "Completion Rate for Primary Education (Female)") +
  theme_minimal() +
  scale_x_discrete(labels = c("Low" = "Low GDP", "Mid" = "Mid GDP", "High" = "High GDP")) +
  coord_cartesian(ylim = c(min(filtered_data$Completion_Rate_Primary_Female), max(filtered_data$Completion_Rate_Primary_Female)))

gdp_comp_rate_female
ggsave("gdp_comp_rate_female.png", gdp_comp_rate_female)
```

## Plot each GDP bin with its corresponding completion rate for primaryrate(male)
```{r}
library(ggplot2)

# Filter out rows where Completion_Rate_Primary_Female is not equal to 0
filtered_data <- FINALgdp_global_education %>%
  filter(Completion_Rate_Primary_Male != 0)

# Create a boxplot with adjustments using the filtered data
gdp_comp_rate_male <- ggplot(filtered_data, aes(x = factor(Rank_of_GDP_per_Capita, levels = c("Low", "Mid", "High")), y = Completion_Rate_Primary_Male)) +
  geom_boxplot(fill = "skyblue", color = "blue", alpha = 0.7, width = 0.5) +
  labs(title = "GDP per Capita Rank vs. Completion Rate for Primary Education (Male)",
       x = "Rank of GDP per Capita",
       y = "Completion Rate for Primary Education (Male)") +
  theme_minimal() +
  scale_x_discrete(labels = c("Low" = "Low GDP", "Mid" = "Mid GDP", "High" = "High GDP")) +
  coord_cartesian(ylim = c(min(filtered_data$Completion_Rate_Primary_Male), max(filtered_data$Completion_Rate_Primary_Male)))

gdp_comp_rate_male
ggsave("gdp_comp_rate_male.png", gdp_comp_rate_male)
```

## Plot GDP and Unemployment rate 
```{r}
library(dplyr)

# Filter out rows where IMF or Unemployment_Rate is 0
filtered_data <- FINALgdp_global_education %>%
  filter(IMF != 0 & Unemployment_Rate != 0)

# Create a bar plot for Unemployment_Rate vs. IMF with non-zero values
gdp_unemployment <- ggplot(filtered_data, aes(x = IMF, y = Unemployment_Rate)) +
  geom_point(stat = "identity", fill = "skyblue", color = "blue", alpha = 0.7) +
  labs(title = "GDP Per Capita vs. Unemployment Rate",
       x = "GDP Per Capita",
       y = "Unemployment Rate") +
  theme_minimal() +
  xlim(5000, max(filtered_data$IMF))  # Setting x-axis limits from 500 to maximum IMF value in filtered data

gdp_unemployment
ggsave("gdp_unemployment.png", gdp_unemployment)
```

## GDP vs OOSR_Primary_Age_Male
```{r}
library(ggplot2)
library(dplyr)

filtered_data <- FINALgdp_global_education %>%
  filter(OOSR_Primary_Age_Male != 0)

gdp_oosr_male <- ggplot(filtered_data, aes(x = IMF, y = OOSR_Primary_Age_Male)) +
  geom_point(stat = "identity", fill = "skyblue", color = "blue", alpha = 0.7) +
  labs(x = "GDP Per Capita", y = "Out-of-school Rate for Primary Age Males",
       title = "GDP Per Capita vs Out-of-school Rate for Primary Age Males") +
  theme_minimal()

gdp_oosr_male
ggsave("gdp_oosr_male.png", gdp_oosr_male)
```

## GDP vs OOSR_Primary_Age_Female
```{r}
library(ggplot2)
library(dplyr)

filtered_data <- FINALgdp_global_education %>%
  filter(OOSR_Primary_Age_Female != 0)

gdp_oosr_female <- ggplot(filtered_data, aes(x = IMF, y = OOSR_Primary_Age_Female)) +
  geom_point(stat = "identity", fill = "skyblue", color = "blue", alpha = 0.7) +
  labs(x = "GDP Per Capita", y = "Out-of-school Rate for Primary Age Females",
       title = "GDP Per Capita vs Out-of-school Rate for Primary Age Females") +
  theme_minimal()

gdp_oosr_female
ggsave("gdp_oosr_female.png", gdp_oosr_female)
```

## New numerical variable: the absolute value of the completion rate of males vs females, so it just shows the difference of males that complete school vs feamles. It shows that in countries w lower GDPs, women have a bigger education gap than men.
```{r}
library(dplyr)

FINALgdp_global_education <- FINALgdp_global_education %>%
  mutate(Education_Gap = abs(Completion_Rate_Primary_Male - Completion_Rate_Primary_Female))

## just to see
ls(FINALgdp_global_education)

## plotting gdp vs educational gap
filtered_data <- FINALgdp_global_education %>%
  filter(!is.na(IMF) & IMF != 0 & !is.na(Education_Gap) & Education_Gap != 0)

# Plotting the scatterplot with filtered data
gdp_education_gap <- ggplot(filtered_data, aes(x = IMF, y = Education_Gap)) +
  geom_point(stat = "identity", fill = "skyblue", color = "blue", alpha = 0.7) + 
  labs(x = "GDP Per Capita", y = "Education Gap") +
  ggtitle("GDP Per Capita vs. Education Gap") +
  theme_minimal()

gdp_education_gap
ggsave("gdp_education_gap.png", gdp_education_gap)
```

## Summary Table
```{r}
summary_table <- FINALgdp_global_education %>%
  group_by(Rank_of_GDP_per_Capita = factor(Rank_of_GDP_per_Capita, levels = c("Low", "Mid", "High", "NA: Insufficient Data"))) %>%
  summarise(
    Median_IMF = median(IMF, na.rm = TRUE),
    Median_Completion_Rate_Primary_Female = median(Completion_Rate_Primary_Female, na.rm = TRUE),
    Median_Completion_Rate_Primary_Male = median(Completion_Rate_Primary_Male, na.rm = TRUE),
    Median_Unemployment_Rate = median(Unemployment_Rate, na.rm = TRUE),
    Median_OOSR_Primary_Age_Male = median(OOSR_Primary_Age_Male, na.rm = TRUE),
    Median_OOSR_Primary_Age_Female = median(OOSR_Primary_Age_Female, na.rm = TRUE),
    Median_Education_Gap = median(Education_Gap, na.rm = TRUE)
  )

# Label NA row as "Insufficient Data"
summary_table$Rank_of_GDP_per_Capita[is.na(summary_table$Rank_of_GDP_per_Capita)] <- "NA: Insufficient Data"

summary_table
write.csv(summary_table, "summary_table.csv", row.names = FALSE)
```

## Map Plot 
```{r}
library(maps)
library(ggplot2)

world_map <- map_data("world")

names(FINALgdp_global_education)[names(FINALgdp_global_education) == "region"] <- "Country.Territory"

merged_data <- merge(world_map, FINALgdp_global_education, by.x = "region", by.y = "Country.Territory")

gdp_map <- ggplot() +
  geom_map(data = merged_data,
           map = world_map,
           aes(x = long, y = lat, map_id = region, fill = Rank_of_GDP_per_Capita),
           color = "black", size = 0.2) +
  scale_fill_manual(values = c("Low" = "red", "Mid" = "yellow", "High" = "green", "NA" = "gray")) +
  theme_minimal() +
  labs(title = "World GDP per Capita Rank",
       subtitle = "Geographical Representation of GDP Per Capita ",
       fill = "GDP per Capita Rank Map") +
  theme_minimal()

gdp_map
ggsave("gdp_map.png", gdp_map, width = 10, height = 5, units = "in")
```

## Counts
```{r}
nrow(FINALgdp_global_education)
ncol(FINALgdp_global_education)

# Counting NAs
na_count <- sapply(FINALgdp_global_education, function(x) sum(is.na(x)))

# Counting zeros
zero_count <- sapply(FINALgdp_global_education, function(x) sum(x == 0))

# Combine counts into a data frame
counts_df <- data.frame(NA_Count = na_count, Zero_Count = zero_count)

counts_df
```
